```{r biasshiehhetunbalSDRN,include=FALSE}
SDNr1 <- NULL #SDNr= sdj^2/nj
SDNr2 <- NULL
NUM <- NULL
DENOM <- NULL
DF <- NULL
coeffmult <- NULL

for (i in 1:100){
  
  n1 <- 50
  n2 <- 20
  N <- n1+n2

  sdrn1 <- i
  sdrn2 <- i
  num <- (sdrn1+sdrn2)^2
  denom <- (sdrn1^2/(n1-1)+sdrn2^2/(n2-1))
  df <- num/denom
  
  SDNr1 <- c(SDNr1,sdrn1)
  SDNr2 <- c(SDNr2,sdrn2)
  NUM <- c(NUM,num)
  DENOM <- c(DENOM,denom)
  DF <- c(DF,df)
  coeffmult <- c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}
```


```{r biasshiehhetunbalSDRN2,fig.cap="Degrees of freedom (DF) when computing the bias of Shieh's $d_s$, when variances and sample sizes are unequal across groups, as a function of the variances and sample sizes ratios ($\\frac{\\sigma_j}{n_j}$)",echo=FALSE}
par(mfrow=c(1,2))
#plot(SDNr2,NUM)
#plot(SDNr2,DENOM)
plot(SDNr1,DF,xlab=expression(paste(sigma[1],"/",n[1])))
plot(SDNr2,DF,xlab=expression(paste(sigma[2],"/",n[2])))
```

```{r biasshiehhetunbalSDRNandnpairingcase1,include=FALSE}
DF <- NULL
coeffmult <- NULL
Nratio <- NULL

for (i in 6:100){

  sdrn1 <- 1
  sdrn2 <- 1
  
  N = 106
  n1 <- i
  n2 <- N-i
  nratio <- n1/n2
  
  df <- (sdrn1+sdrn2)^2/(sdrn1^2/(n1-1)+sdrn2^2/(n2-1))
 
  Nratio <- c(Nratio,nratio)  
  DF <- c(DF,df)
  coeffmult <- c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}
```


```{r biasshiehhetunbalSDRNandnpairing2case1,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Shieh's $d_s$, when variances and sample sizes are unequal across groups and $\\frac{\\sigma^2_1}{n_1}=\\frac{\\sigma^2_2}{n_2}$, as a function of the logarithm of the sample sizes ratio ($log \\left( \\frac{n_1}{n_2} \\right)$)",echo=FALSE}
par(mfrow=c(1,2))
#the plot(SDNr2,NUM)
#plot(SDNr2,DENOM)
plot(log(Nratio),DF,xlab=expression(paste("log(",n[1],"/",n[2],")")))
#Nratio[DF==max(DF)]
plot(log(Nratio),coeffmult,ylab=expression(gamma),xlab=expression(paste("log(",n[1],"/",n[2],")")))
#Nratio[coeffmult==min(coeffmult)]
```

```{r biasshiehhetunbalSDRNandnpairingcase2,include=FALSE}
Nratio1 <- NULL
DF1 <- NULL
coeffmult1 <- NULL

for (i in 6:100){

  sdrn1 <- 1
  sdrn2 <- 2
  
  N = 106
  n1 <- i
  n2 <- N-i
  nratio <- n1/n2
  
  df <- (sdrn1+sdrn2)^2/(sdrn1^2/(n1-1)+sdrn2^2/(n2-1))
 
  Nratio1 <- c(Nratio1,nratio)  
  DF1 <- c(DF1,df)
  coeffmult1 <- c(coeffmult1,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}

#----------------------------------------------------------------------

Nratio2 <- NULL
DF2 <- NULL
coeffmult2 <- NULL

for (i in 6:100){

  sdrn1 <- 1
  sdrn2 <- 10
  
  N = 106
  n1 <- i
  n2 <- N-i
  nratio <- n1/n2
  
  df <- (sdrn1+sdrn2)^2/(sdrn1^2/(n1-1)+sdrn2^2/(n2-1))
 
  Nratio2 <- c(Nratio2,nratio)  
  DF2 <- c(DF2,df)
  coeffmult2 <- c(coeffmult2,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}

#----------------------------------------------------------------------

Nratio3 <- NULL
DF3 <- NULL
coeffmult3 <- NULL

for (i in 6:100){

  sdrn1 <- 2
  sdrn2 <- 1
  
  N = 106
  n1 <- i
  n2 <- N-i
  nratio <- n1/n2
  
  df <- (sdrn1+sdrn2)^2/(sdrn1^2/(n1-1)+sdrn2^2/(n2-1))
 
  Nratio3 <- c(Nratio3,nratio)  
  DF3 <- c(DF3,df)
  coeffmult3 <- c(coeffmult3,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}

#----------------------------------------------------------------------

Nratio4 <- NULL
DF4 <- NULL
coeffmult4 <- NULL

for (i in 6:100){

  sdrn1 <- 10
  sdrn2 <- 1
  
  N = 106
  n1 <- i
  n2 <- N-i
  nratio <- n1/n2
  
  df <- (sdrn1+sdrn2)^2/(sdrn1^2/(n1-1)+sdrn2^2/(n2-1))
 
  Nratio4 <- c(Nratio4,nratio)  
  DF4 <- c(DF4,df)
  coeffmult4 <- c(coeffmult4,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}

```


```{r biasshiehhetunbalSDRNandnpairing2case2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Shieh's $d_s$, when variances and sample sizes are unequal across groups and either $\\frac{\\sigma^2_1}{n_1}>\\frac{\\sigma^2_2}{n_2}$ (top) or $\\frac{\\sigma^2_1}{n_1}<\\frac{\\sigma^2_2}{n_2}$ (bottom), as a function of the logarithm of the sample sizes ratio ($log \\left( \\frac{n_1}{n_2} \\right)$)",echo=FALSE}
par(mfrow=c(2,2))
plot(log(Nratio3),coeffmult3,main="when SDNratio1=2 and SDNratio1=1",ylab=expression(gamma),xlab=expression(paste("log(",n[1],"/",n[2],")")))
abline(v=log(Nratio3)[coeffmult3==min(coeffmult3)])
plot(log(Nratio4),coeffmult4,main="when SDNratio1=10 and SDNratio1=1",ylab=expression(gamma),xlab=expression(paste("log(",n[1],"/",n[2],")")))
abline(v=log(Nratio4)[coeffmult4==min(coeffmult4)])
plot(log(Nratio1),coeffmult1,main="when SDNratio1=1 and SDNratio1=2",ylab=expression(gamma),xlab=expression(paste("log(",n[1],"/",n[2],")")))
abline(v=log(Nratio1)[coeffmult1==min(coeffmult1)])
plot(log(Nratio2),coeffmult2,main="when SDNratio1=1 and SDNratio1=10",ylab=expression(gamma),xlab=expression(paste("log(",n[1],"/",n[2],")")))
abline(v=log(Nratio2)[coeffmult2==min(coeffmult2)])
```

  + The variances and sample sizes ratios don't matter per se (see Figure \ref{fig:biasshiehhetunbalSDRN2}). However, the pairing between these ratios and sample sizes has an effect on the bias:
    - When $\frac{\sigma^2_1}{n_1}=\frac{\sigma^2_2}{n_2}$, the smallest bias occurs when sample sizes are equal across groups. The further the sample sizes ratio is from 1, the larger the bias (see Figure \ref{fig:biasshiehhetunbalSDRNandnpairing2case1}).
    - When $\frac{\sigma^2_1}{n_1} \neq \frac{\sigma^2_2}{n_2}$, the minimum bias will always occure when $min(\frac{\sigma^2_j}{n_j})$ will be associated with $min(n_j)$. In other word, when $\frac{\sigma^2_1}{n_1}>\frac{\sigma^2_2}{n_2}$, the sample sizes ratio associated with the minimum bias will be positive, meaning that $n_1>n_2$ (and the larger the difference between $\frac{\sigma^2_1}{n_1}$ and $\frac{\sigma^2_2}{n_2}$, the further from 1 will be this sample sizes ratio; see the two top plots in Figure \ref{fig:biasshiehhetunbalSDRNandnpairing2case2}). On the other side, when $\frac{\sigma^2_1}{n_1}<\frac{\sigma^2_2}{n_2}$, the sample sizes ratio associated with the minimum bias will be negative, meaning that $n_1<n_2$ (and the larger the difference between $\frac{\sigma^2_1}{n_1}$ and $\frac{\sigma^2_2}{n_2}$, the further from 1 will be this sample sizes ratio; see the two bottom plots in Figure \ref{fig:biasshiehhetunbalSDRNandnpairing2case2}). 
