---
title             : "Theoretical Bias, as a function of population parameters"
shorttitle        : "Theoretical Bias"

author: 
  - name          : "Marie Delacre" 
    affiliation   : "1"
    corresponding : yes    
    address       : "CP191, avenue F.D. Roosevelt 50, 1050 Bruxelles"
    email         : "marie.delacre@ulb.be"
  - name          : "Daniel Lakens"
    affiliation   : "2"
  - name          : "Christophe Ley"
    affiliation   : "3"
  - name          : "Limin Liu"
    affiliation   : "3"
  - name          : "Christophe Leys"
    affiliation   : "1"
    
affiliation:
  - id            : "1"
    institution   : "Université Libre de Bruxelles, Service of Analysis of the Data (SAD), Bruxelles, Belgium"
  - id            : "2"
    institution   : "Eindhoven University of Technology, Human Technology Interaction Group, Eindhoven, the Netherlands"
  - id            : "3"
    institution   : "Universiteit Gent, Department of Applied Mathematics, Computer Science and Statistics, Gent, Belgium"

wordcount         : "2287 words"

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

# The bias 

For all "biased" estimators, when the population effect size is null so is the bias. We will therefore focus on configurations where there is a non-null population effect size. The sampling distribution of Cohen's $d_s$ (and therefore its bias) is only known under the assumptions of normality and homoscedasticity. On the other hand, the biases of Glass's $d_s$, Cohen's $d^*_s$ and Shieh's $d_s$ are theoretically known for all configurations where the normality assumption is met. In order to simplify the analysis of their bias, it is convenient to subdivide all configurations into 3 conditions:  
- when population variances are equal across groups;     
- when population variances are unequal across groups, with equal sample sizes;  
- when population variances are unequal across groups, with unequal sample sizes.  

## Preliminary note

For all previously mentioned estimators (Cohen's $d_s$, Glass's $d_s$, Cohen's $d^*_s$ and Shieh's $d_s$), the theoretical expectation is computed by multiplying the population effect size (respectively $\delta_{Cohen}$, $\delta_{Glass}$, $\delta^*_{Cohen}$ and $\delta_{Shieh}$) by the following multiplier coefficient:
\begin{equation} 
\gamma=\frac{\sqrt{\frac{df}{2}} \times \Gamma{\frac{df-1}{2}}}{\Gamma{\frac{df}{2}}}
(\#eq:mc)
\end{equation} 
where $df$ are the degrees of freedom (see the main article). $\gamma$ is *always* positive, meaning that when the population effect size is not zero, all estimators will overestimate the population effect size. Moreover, its limit tends to 1 when the degrees of freedom ($df$) tend to infinity, meaning that the larger the degrees of freedom, the lower the bias.   
While we focus on the theoretical bias of biased estimators when the normality assumption is met, it is interesting to notice that our main conclusions seem to generalize to :  
- biased estimators when samples are extracted from symmetric distributions;     
- unbiased estimators when samples are extracted from heavy-tailed symmetric distributions.  

## Cohen's $\bf d_s$ (see Table 1)

Under the assumptions that independant residuals are normally distributed with equal variances, the **bias** of Cohen's $d_s$ is a function of total sample size ($N$) and the population effect size ($\delta_{Cohen}$):     

  + The larger the population effect size, the more Cohen's $d_s$ will overestimate $\delta_{Cohen}$;   
  
```{r biascohendNsize,include=FALSE}
Nsize=NULL
coeffmult=NULL
DF=NULL

sd1=2
sd2=2

for (i in 4:200){
  n1=i
  n2=i
  N = n1+n2
  Nsize=c(Nsize,N)
  df = n1+n2-2
  DF = c(DF,df)
  coeffmult = c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}
```

```{r biascohendNsize2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Cohen's $d_s$, when variances are equal across groups, as a function of the total sample size ($N$).",echo=FALSE}
par(mfrow=c(1,2))
plot(Nsize,DF,xlab="N")
plot(Nsize,coeffmult,ylab=expression(gamma),xlab="N")
```

  + The larger the total sample size, the lower the bias (see Figure \ref{fig:biascohendNsize2});

  + Of course, considering the degrees of freedom, the sample size ratio does not matter (i.e. the bias will decrease when increasing $n_1$, $n_2$ or both sample sizes).

## Glass's $\bf d_s$ (see Table 2)

Because degrees of freedom depend only on the control group size (neither on $\sigma_1$ nor on $\sigma_2$), there is no need to distinguish between cases where there is homoscedasticity or heteroscedasticity!   

The **bias** of Glass's $d_s$ is a function of the control group size ($n_c$) and the population effect size ($\delta_{Glass}$):  

  + The larger the population effect size, the more Glass's $d_s$ will overestimate $\delta_{Glass}$;

```{r biasGlassctrlsize,include=FALSE}
Nctrl=NULL
coeffmult=NULL
DF=NULL

sd1=2
sd2=2

for (i in 4:200){
  n1=i
  n2=20
  Nctrl=c(Nctrl,n1)
  df = n1-2
  DF = c(DF,df)
  coeffmult = c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}

Nexp=NULL
coeffmult2=NULL
DF2=NULL

sd1=2
sd2=2

for (i in 4:200){
  n1=20
  n2=i
  Nexp=c(Nexp,n2)
  df = n1-2
  DF2 = c(DF2,df)
  coeffmult2 = c(coeffmult2,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}

```

```{r biasGlassctrlsize2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Glass's $d_s$, when variances are equal across groups, as a function of $n_c$ (top) and $n_e$ (bottom).",echo=FALSE}
par(mfrow=c(2,2))
plot(Nctrl,DF,xlab=expression(n[c]))
plot(Nctrl,coeffmult,ylab=expression(gamma),xlab=expression(n[c]))
plot(Nexp,DF2,xlab=expression(n[e]))
plot(Nexp,coeffmult2,ylab=expression(gamma),xlab=expression(n[e]))
```

  + The larger the size of the control group, the lower the bias (see the two top plots in Figure \ref{fig:biasGlassctrlsize2}). On the other hand, increasing the experimental group size does not impact the bias (see the two bottom plots in Figure \ref{fig:biasGlassctrlsize2}).

## Cohen's $\bf d^*_s$ (see Table 2)

### When variances are equal across populations

When $\sigma_1=\sigma_2=\sigma$: 
$$df_{Cohen's \; d^*_s} = \frac{(n_1-1)(n_2-1)(2\sigma^2)^2}{(n_2-1)\sigma^4+(n_1-1)\sigma^4} = \frac{(n_1-1)(n_2-1)\times 4\sigma^4}{\sigma^4(n_1+n_2-2)} = \frac{4(n_1-1)(n_2-1)}{n_1+n_2-2}$$ 
One can see that degrees of freedom depend only on the total sample size ($N$) and the sample size allocation ratio $\left( \frac{n_1}{n_2}\right)$. As a consequence, the **bias** of Cohen's $d^*_s$ is a function of the population effect size ($\delta^*_{Cohen}$), the sample size allocation ratio $\left( \frac{n_1}{n_2}\right)$ and the total sample size ($N$). 

  + The larger the population effect size, the more Cohen's $d^*_s$ will overestimate $\delta^*_{Cohen}$;

```{r biascohendprimehomNratio,include=FALSE}
coeffmult <- NULL
nratio <- NULL
DF <- NULL

for (i in 10:190){
  
  N <- 200
  n1 <- i
  n2 <- N-n1
  sd1 <- 12
  sd2 <- 12
  
  nratio <- c(nratio,n1/n2)
  df <- ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)/((n2-1)*sd1^4+(n1-1)*sd2^4)
  DF <- c(DF,df)
  coeffmult <- c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
  
}
```

```{r biascohendprimehomNratio2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Cohen's $d^*_s$, when variances are equal across groups, as a function of the logarithm of the sample sizes ratio $log\\left(\\frac{n_1}{n_2} \\right)$.",echo=FALSE}
par(mfrow=c(1,2))
plot(log(nratio),DF,xlab=expression(paste("log(",n[1],"/",n[2],")")))
#nratio[DF==max(DF)]
plot(log(nratio),coeffmult,ylim=c(1,1.02),ylab=expression(gamma),xlab=expression(paste("log(",n[1],"/",n[2],")")))
#nratio[coeffmult==min(coeffmult)]
```

  + The further the sample size allocation ratio is from 1, the larger the bias (see Figure \ref{fig:biascohendprimehomNratio2});

```{r biascohendprimehomNsize,include=FALSE}
Nsize=NULL
coeffmult=NULL
DF=NULL

sd1=2
sd2=2

for (i in 4:200){
  n1=i
  n2=10
  N = n1+n2
  Nsize=c(Nsize,N)
  df = ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)/((n2-1)*sd1^4+(n1-1)*sd2^4) 
  DF = c(DF,df)
  coeffmult = c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}
```

```{r biascohendprimehomNsize2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Cohen's $d^*_s$, when variances are equal across groups, as a function of the total sample size ($N$).",echo=FALSE}
par(mfrow=c(1,2))
plot(Nsize,DF,xlab="N")
plot(Nsize,coeffmult,ylab=expression(gamma),xlab="N")
```

  + The larger the total sample size, the lower the bias (see Figure \ref{fig:biascohendprimehomNsize2}).

### When variances are unequal across populations, with equal sample sizes
When $n_1 = n_2 = n$:
$$df_{Cohen's \; d^*_s} = \frac{(n-1)^2(\sigma^2_1+\sigma^2_2)^2}{(n-1)(\sigma^4_1+\sigma^4_2)} =  \frac{(n-1)(\sigma^4_1+\sigma^4_2+2\sigma^2_1\sigma^2_2)}{\sigma^4_1+\sigma^4_2}$$ 
One can see that degrees of freedom depend only on the total sample size  ($N$) and the $SD$-ratio $\left( \frac{\sigma_1}{\sigma_2}\right)$. As a consequence, the **bias** of Cohen's $d^*_s$ is a function of the population effect size ($\delta^*_{Cohen}$), the $SD$-ratio $\left( \frac{\sigma_1}{\sigma_2}\right)$ and the total sample size ($N$): 

  + The larger the population effect size, the more Cohen's $d^*_s$ will overestimate $\delta^*_{Cohen}$;  

```{r biascohendprimehetbalSDratio,include=FALSE}
coeffmult <- NULL
SDratio <- NULL
DF <- NULL

for (i in 1:100){
  
  n1=100
  n2=100
  N <- n1+n2
  sd2 <- 10
  sd1 <- i
  
  SDratio <- c(SDratio,sd1/sd2)
  df <- ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)/((n2-1)*sd1^4+(n1-1)*sd2^4)
  DF <- c(DF,df)
  coeffmult <- c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}
```

```{r biascohendprimehetbalSDratio2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Cohen's $d^*_s$, when variances are unequal across groups and sample sizes are equal, as a function of the logarithm of the $SD$-ratio ($log \\left( \\frac{\\sigma_1}{\\sigma_2} \\right)$).",echo=FALSE}
par(mfrow=c(1,2))
plot(log(SDratio),DF,,xlab=expression(paste("log(",sigma[1],"/",sigma[2],")")))
#SDratio[DF==max(DF)]
plot(log(SDratio),coeffmult,ylab=expression(gamma),xlab=expression(paste("log(",sigma[1],"/",sigma[2],")")))
#SDratio[coeffmult==min(coeffmult)]
```

  + The further the $SD$-ratio is from 1, the larger the bias (see Figure \ref{fig:biascohendprimehetbalSDratio2});  

```{r biascohendprimehetbalNsize,include=FALSE}
Nsize=NULL
coeffmult=NULL
DF=NULL

sd1=10
sd2=2

for (i in 4:200){
  n1=i
  n2=i
  N = n1+n2
  Nsize=c(Nsize,N)
  df = ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)/((n2-1)*sd1^4+(n1-1)*sd2^4) 
  DF = c(DF,df)
  coeffmult = c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}
```

```{r biascohendprimehetbalNsize2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Cohen's $d^*_s$, when variances are unequal across groups and sample sizes are equal, as a function of the total sample size ($N$).",echo=FALSE}
par(mfrow=c(1,2))
plot(Nsize,DF,xlab="N")
plot(Nsize,coeffmult,ylab=expression(gamma),xlab="N")
```

  + The larger the total sample size, the lower the bias (see Figure \ref{fig:biascohendprimehetbalNsize2}).  

```{r biascohendprimehetbalvariance,include=FALSE}
coeffmult <- NULL
SD1 <- NULL
SD2 <- NULL
DF <- NULL

for (i in 1:100){
  n1=100
  n2=100
  N <- n1+n2
  sdratio = 1/2
  sd1 <- i
  sd2 <- sdratio*i

  SD1 <- c(SD1,sd1)
  SD2 <- c(SD2,sd2)
  df <- ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)/((n2-1)*sd1^4+(n1-1)*sd2^4)
  DF <- c(DF,df)
  coeffmult <- c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}
```

```{r biascohendprimehetbalvariance2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Cohen's $d^*_s$, when variances are unequal across groups and sample sizes are equal, as a function of $\\sigma_1$ (top plots) and $\\sigma_2$ (bottom plots), for a constant $SD$-ratio.",echo=FALSE}
par(mfrow=c(2,2))
plot(SD1,DF,xlab=expression(sigma[1]))
plot(SD1,coeffmult,ylab=expression(gamma),xlab=expression(sigma[1]))

plot(SD2,DF,xlab=expression(sigma[2]))
plot(SD2,coeffmult,ylab=expression(gamma),xlab=expression(sigma[2]))
```

Note: for a constant $SD$-ratio, $\sigma_1$ and $\sigma_2$ do not matter (see Figure \ref{fig:biascohendprimehetbalvariance2}). 

### When variances are unequal across populations, with unequal sample sizes

The **bias** of Cohen's $d^*_s$ is a function of the population effect size ($\delta^*_{Cohen}$), the total sample size ($N$), and the interaction between the sample sizes ratio and the $SD$-ratio $\left(\frac{n_1}{n_2}\times\frac{\sigma_1}{\sigma_2} \right)$ :     

  + The larger the population effect size, the more Cohen's $d^*_s$ will overestimate $\delta^*_{Cohen}$;  

```{r biascohendprimehetunbalNsize,include=FALSE}
coeffmult <- NULL
Nsize <- NULL
DF <- NULL

for (i in 2:200){
  
  n1 <- i
  n2 <- 12*i
  N <- n1+n2
  sd1 <- 1.8
  sd2 <- 1
  
  df <- ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)/((n2-1)*sd1^4+(n1-1)*sd2^4)
  DF <- c(DF,df)
  Nsize <- c(Nsize,N)
  coeffmult <- c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}

```

```{r biascohendprimehetunbalNsize2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Cohen's $d^*_s$, when variances and sample sizes are unequal across groups, as a function of the total sample size ($N$).",echo=FALSE}
par(mfrow=c(1,2))
plot(Nsize,DF,xlab="N")
plot(Nsize,coeffmult,ylab=expression(gamma),xlab="N")
```

  + The larger the total sample size, the lower the bias (see Figure \ref{fig:biascohendprimehetunbalNsize2});

```{r biascohendprimehetunbalnratiosdratio,include=FALSE}
coeffmult2 <- NULL
nratio2 <- NULL
DF2 <- NULL
DF_NUM2 <- NULL
DF_DENOM2 <- NULL

for (i in 10:190){
  
  delta_cohendprime <- 0
  N <- 200
  n1 <- i
  n2 <- N-n1
  sd2 <- 1
  sd1 <- sqrt(50-sd2^2) # so sqrt((sd1^2+sd2^2)/2)=5
  SDratio2=sd1/sd2
  nratio2 <- c(nratio2,n1/n2)
  df_num <- ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)
  df_denom <- ((n2-1)*sd1^4+(n1-1)*sd2^4)
  DF_NUM2 <- c(DF_NUM2, df_num)
  DF_DENOM2 <- c(DF_DENOM2, df_denom)
  df <- ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)/((n2-1)*sd1^4+(n1-1)*sd2^4)
  DF2 <- c(DF2,df)
  coeffmult2 <- c(coeffmult2,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}

coeffmult3 <- NULL
nratio3 <- NULL
DF3 <- NULL
DF_NUM3 <- NULL
DF_DENOM3 <- NULL

for (i in 10:190){

  delta_cohendprime <- 0  
  N <- 200
  n1 <- i
  n2 <- N-n1
  sd2 <- 2
  sd1 <- sqrt(50-sd2^2) # so sqrt((sd1^2+sd2^2)/2)=5
  SDratio3=sd1/sd2
  nratio3 <- c(nratio3,n1/n2)
  df_num <- ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)
  df_denom <- ((n2-1)*sd1^4+(n1-1)*sd2^4)
  DF_NUM3 <- c(DF_NUM3, df_num)
  DF_DENOM3 <- c(DF_DENOM3, df_denom)
  df <- ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)/((n2-1)*sd1^4+(n1-1)*sd2^4)
  DF3 <- c(DF3,df)
  coeffmult3 <- c(coeffmult3,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}

coeffmult4 <- NULL
nratio4 <- NULL
DF4 <- NULL
DF_NUM4 <- NULL
DF_DENOM4 <- NULL

for (i in 10:190){

  delta_cohendprime <- 0  
  N <- 200
  n1 <- i
  n2 <- N-n1
  sd2 <- 4
  sd1 <- sqrt(50-sd2^2) # so sqrt((sd1^2+sd2^2)/2)=5
  SDratio4=sd1/sd2
  nratio4 <- c(nratio4,n1/n2)
  df_num <- ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)
  df_denom <- ((n2-1)*sd1^4+(n1-1)*sd2^4)
  DF_NUM4 <- c(DF_NUM4, df_num)
  DF_DENOM4 <- c(DF_DENOM4, df_denom)
  df <- ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)/((n2-1)*sd1^4+(n1-1)*sd2^4)
  DF4 <- c(DF4,df)
  coeffmult4 <- c(coeffmult4,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}
```

```{r biascohendprimehetunbalnratiosdratio2,fig.cap="Degrees of freedom ($DF$) and $\\gamma$, when computing the bias of Cohen's $d^*_s$ when variances and sample sizes are unequal across groups, as a function of the logarithm of the sample sizes ratio ($log \\left( \\frac{n_1}{n_2} \\right)$), when $SD$-ratio equals 1.46 (first row), 3.39 (second row) or 7 (third row).",echo=FALSE}
par(mfrow=c(3,2))

plot(log(nratio4),DF4,ylab="DF",xlab=expression(paste("log(",n[1],"/",n[2],")")))
plot(log(nratio4),coeffmult4,ylab=expression(gamma),xlab=expression(paste("log(",n[1],"/",n[2],")")))

plot(log(nratio3),DF3,ylab="DF",xlab=expression(paste("log(",n[1],"/",n[2],")")))
plot(log(nratio3),coeffmult3,ylab=expression(gamma),xlab=expression(paste("log(",n[1],"/",n[2],")")))

plot(log(nratio2),DF2,ylab="DF",xlab=expression(paste("log(",n[1],"/",n[2],")")))
plot(log(nratio2),coeffmult2,ylab=expression(gamma),xlab=expression(paste("log(",n[1],"/",n[2],")")))
```
 
  + The smallest bias always occurs when there is a positive pairing between variances and sample sizes, because one gives more weight to the smallest variance, in the denominator of the $df$ computation. Moreover, the further the $SD$-ratio is from 1, the further from 1 will also be the sample sizes ratio associated with the smallest bias (see Figure \ref{fig:biascohendprimehetunbalnratiosdratio2}). This can be explained by splitting the numerator and the denominator in the $df$ computation.
  
```{r dfnumdenomcohendprimehetunbalnratiosdratio2,fig.cap="Numerator and denominator of the degrees of freedom ($DF$) computation, when computing the bias of Cohen's $d^*_s$ when variances and sample sizes are unequal across groups, as a function of the logarithm of the sample sizes ratio ($log \\left( \\frac{n_1}{n_2} \\right)$), when $SD$-ratio equals 1.46 (first row), 3.39 (second row) or 7 (third row).",echo=FALSE}
par(mfrow=c(3,2),mar=c(4,3,1,3))

plot(log(nratio4),DF_NUM4,ylab="DF",xlab=expression(paste("log(",n[1],"/",n[2],")")))
plot(log(nratio4),DF_DENOM4,ylab=expression(gamma),xlab=expression(paste("log(",n[1],"/",n[2],")")))

plot(log(nratio3),DF_NUM3,ylab="DF",xlab=expression(paste("log(",n[1],"/",n[2],")")))
plot(log(nratio3),DF_DENOM3,ylab=expression(gamma),xlab=expression(paste("log(",n[1],"/",n[2],")")))

plot(log(nratio2),DF_NUM2,ylab="DF",xlab=expression(paste("log(",n[1],"/",n[2],")")))
plot(log(nratio2),DF_DENOM2,ylab=expression(gamma),xlab=expression(paste("log(",n[1],"/",n[2],")")))
```

As illustrated in Figure \ref{fig:dfnumdenomcohendprimehetunbalnratiosdratio2}, for any $SD$-ratio, the numerator of the degrees of freedom will be maximized when sample sizes are equal across groups (and is not impacted by the $SD$-ratio). On the other hand, the denominator will be minimized when there is a positive pairing between variances and sample sizes. For example, when $\sigma_1 > \sigma_2$, the smallest denominator occurs when $\frac{n_1}{n_2}$ reaches its maximum value and the larger the $SD$-ratio, the larger the impact of the sample sizes ratio on the denominator. 

```{r biascohendprimehetunbalvariance,include=FALSE}
coeffmult <- NULL
SD <- NULL
DF <- NULL

for (i in 2:200){
  
  n1 <- 23
  n2 <- 75
  N <- n1+n2
  sd1 <- i
  sd2 <- 8*i
  
  df <- ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)/((n2-1)*sd1^4+(n1-1)*sd2^4)
  DF <- c(DF,df)
  SD <- c(SD,sqrt((sd1^2+sd2^2)/2))
  coeffmult <- c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}

```

```{r biascohendprimehetunbalvariance2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Cohen's $d^*_s$, when variances and sample sizes are unequal across groups, as a function of $\\sigma= \\frac{(\\sigma_1^2+\\sigma_2^2)}{2}$, for a constant $SD$-ratio.",echo=FALSE}
par(mfrow=c(2,2))
plot(SD,DF,xlab=expression(sigma))
plot(SD,coeffmult,ylab=expression(gamma),xlab=expression(sigma))
```
  
Note: for a constant $SD$-ratio, the variance does not matter. (See Figure \ref{fig:biascohendprimehetunbalvariance2}).

## Shieh's $\bf d_s$ (see Table 2)

### When variances are equal across populations

When $\sigma_1=\sigma_2=\sigma$:
$$df_{Shieh's \; d_s} = \frac{\left( \frac{n_2\sigma^2+n_1\sigma^2}{n_1n_2}\right)^2}{\frac{(n_2-1)\left( \frac{\sigma^2}{n_1}\right)^2+(n_1-1)\left( \frac{\sigma^2}{n_2}\right)^2}{(n_1-1)(n_2-1)}}$$
$$\leftrightarrow df_{Shieh's \; d_s} = \frac{[\sigma^2(n_1+n_2)]^2}{n_1^2n_2^2} \times \frac{(n_1-1)(n_2-1)}{(n_2-1) \times  \frac{\sigma^4}{n_1^2}+(n_1-1) \times \frac{\sigma^4}{n_2^2}}$$
$$\leftrightarrow df_{Shieh's \; d_s} = \frac{\sigma^4N^2}{n_1^2n_2^2} \times \frac{(n_1-1)(n_2-1)}{\sigma^4 \left( \frac{n_2-1}{n^2_1}+\frac{n_1-1}{n^2_2}\right) }$$
$$\leftrightarrow df_{Shieh's \; d_s} = \frac{N^2(n_1-1)(n_2-1)}{n_1^2n_2^2 \left( \frac{n_2^2(n_2-1)+n_1^2(n_1-1)}{n_1^2n_2^2}\right)}$$
$$\leftrightarrow df_{Shieh's \; d_s} = \frac{N^2(n_1-1)(n_2-1)}{n_2^2(n_2-1)+n_1^2(n_1-1)}$$

One can see that degrees of freedom depend only on the total sample size ($N$) and the sample size allocation ratio $\left( \frac{n_1}{n_2}\right)$. As a consequence, the **bias** of Shieh's $d^*_s$ is a function of the population effect size ($\delta_{Shieh}$), the sample size allocation ratio $\left( \frac{n_1}{n_2}\right)$ and the total sample size ($N$). 

  + The larger the population effect size, the more Shieh's $d_s$ will overestimate $\delta_{Shieh}$;
  
```{r biasshiehhomNratio,include=FALSE}
coeffmult <- NULL
nratio <- NULL
DF <- NULL

for (i in 10:190){
  
  N <- 200
  n1 <- i
  n2 <- N-n1
  sd1 <- 8
  sd2 <- 8
  
  nratio <- c(nratio,n1/n2)
  df <- (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF <- c(DF,df)
  coeffmult <- c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
  
}
```

```{r biasshiehhomNratio2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Shieh's $d_s$, when variances are equal across groups, as a function of the logarithm of the sample sizes ratio $(log \\left(\\frac{n_1}{n_2})\\right)$.",echo=FALSE}
par(mfrow=c(1,2))
plot(log(nratio),DF,xlab=expression(paste("log(",n[1],"/",n[2],")")))
#nratio[DF==max(DF)]
plot(log(nratio),coeffmult,ylim=c(1,1.02),ylab=expression(gamma),xlab=expression(paste("log(",n[1],"/",n[2],")")))
#nratio[coeffmult==min(coeffmult)]
```

  + The further the sample size allocation ratio is from 1, the larger the bias (see Figure \ref{fig:biasshiehhomNratio2});

```{r biasshiehhomNsize,include=FALSE}
Nsize=NULL
coeffmult=NULL
DF=NULL

sd1=2
sd2=2

for (i in 4:200){
  n1=2.45*i
  n2=i
  N = n1+n2
  Nsize=c(Nsize,N)
  df = (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF = c(DF,df)
  coeffmult = c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}
```

```{r biasshiehhomNsize2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Shieh's $d_s$, when variances are equal across groups, as a function of the total sample size ($N$).",echo=FALSE}
par(mfrow=c(1,2))
plot(Nsize,DF,xlab="N")
plot(Nsize,coeffmult,ylab=expression(gamma),xlab="N")
```

  + For a constant sample sizes ratio, the larger the total sample size, the lower the bias (see Figure \ref{fig:biasshiehhomNsize2}).

```{r biasshiehhomuneqNsize,include=FALSE}
Nsize=NULL
coeffmult=NULL
DF=NULL

sd1=2
sd2=2

for (i in 4:200){
  n1=13
  n2=i
  N = n1+n2
  Nsize=c(Nsize,N)
  df = (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF = c(DF,df)
  coeffmult = c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}

Nsize2=NULL
coeffmult2=NULL
DF2=NULL

sd1=2
sd2=2

for (i in 4:200){
  n1=i
  n2=13
  N = n1+n2
  Nsize2=c(Nsize2,N)
  df = (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF2 = c(DF2,df)
  coeffmult2 = c(coeffmult2,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}

```

```{r biasshiehhomuneqNsize2,fig.cap="Degrees of freedom (DF), when computing the bias of Shieh's $d_s$, when variances are equal across groups, when adding subjects only in the first group (left) or in the second group (right).",echo=FALSE}
par(mfrow=c(1,2))
plot(Nsize,DF,xlab=expression(n[1]))
plot(Nsize2,DF2,ylab="DF",xlab=expression(n[2]))
```

Note: when computing Cohen's $d^*_s$, degrees of freedom increased when adding subjects in either one or both groups, even when the sample size ratio increased. When computing Shieh's $d_s$, this is not true anymore: there is a larger impact of the sample sizes ratio such that increasing the sample sizes ratio when adding subjects in only one group can decrease the degrees of freedom and therefore, increase the bias (See Figure \ref{fig:biasshiehhomuneqNsize2}).

### When variances are unequal across populations, with equal sample sizes

When $n_1=n_2=n$:
$$df_{Shieh's \; d_s} = \frac{\left( \frac{\sigma_1^2+\sigma_2^2}{n} \right)^2}{\frac{(\sigma_1^2/n)^2+(\sigma_2^2/n)^2}{n-1}}$$
$$\leftrightarrow df_{Shieh's \; d_s} = \frac{(\sigma_1^2+\sigma_2^2)^2}{n^2} \times\frac{n-1}{\frac{\sigma_1^4+\sigma_2^4}{n^2}}$$
$$\leftrightarrow df_{Shieh's \; d_s} = \frac{(\sigma_1^2+\sigma_2^2)^2 \times (n-1)}{\sigma_1^4+\sigma_2^4}$$

One can see that degrees of freedom depend on the total sample size ($N$) and the $SD$-ratio $\left( \frac{\sigma_1}{\sigma_2}\right)$.  As a consequence, the bias depends on the population effect size ($\delta_{Shieh}$), the $SD$-ratio $\left( \frac{\sigma_1}{\sigma_2}\right)$ and the total sample size ($N$).

  + The larger the population effect size, the more Shieh's $d_s$ will overestimate $\delta_{Shieh}$;  

```{r biasshiehhetbalSDratio,include=FALSE}
coeffmult <- NULL
SDratio <- NULL
DF <- NULL

for (i in 1:100){
  
  n1=100
  n2=100
  N <- n1+n2
  sd2 <- 10
  sd1 <- i
  
  SDratio <- c(SDratio,sd1/sd2)
  df <- (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF <- c(DF,df)
  coeffmult <- c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}
```

```{r biasshiehhetbalSDratio2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Shieh's $d_s$, when variances are unequal across groups and sample sizes are equal, as a function of the logarithm of the $SD$-ratio $(log \\left(\\frac{\\sigma_1}{\\sigma_2})\\right)$.",echo=FALSE}
par(mfrow=c(1,2))
plot(log(SDratio),DF,xlab=expression(paste("log(",sigma[1],"/",sigma[2],")")))
#SDratio[DF==max(DF)]
plot(log(SDratio),coeffmult,ylab=expression(gamma),xlab=expression(paste("log(",sigma[1],"/",sigma[2],")")))
#SDratio[coeffmult==min(coeffmult)]
```

  + The further the $SD$-ratio is from 1, the larger the bias (see Figure \ref{fig:biasshiehhetbalSDratio2});  

```{r biasshiehhetbalNsize,include=FALSE}
Nsize=NULL
coeffmult=NULL
DF=NULL

sd1=12
sd2=1.4786

for (i in 4:200){
  n1=i
  n2=i
  N = n1+n2
  Nsize=c(Nsize,N)
  df = (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1)) 
  DF = c(DF,df)
  coeffmult = c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}
```

```{r biasshiehhetbalNsize2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Shieh's $d_s$, when variances are unequal across groups and sample sizes are equal, as a function of the total sample size ($N$).",echo=FALSE}
par(mfrow=c(1,2))
plot(Nsize,DF,xlab="N")
plot(Nsize,coeffmult,ylab=expression(gamma),xlab="N")
```

  + The larger the total sample size, the lower the bias (see Figure \ref{fig:biasshiehhetbalNsize2});  

```{r biasshiehhetbalvariance,include=FALSE}
coeffmult <- NULL
SD1 <- NULL
DF <- NULL

for (i in 1:200){
  n1=100
  n2=100
  N <- n1+n2
  sdratio = 3
  sd1 <- i
  sd2 <- sdratio*i

  SD1 <- c(SD1,sd1)
  df <- (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF <- c(DF,df)
  coeffmult <- c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}
```

```{r biasshiehhetbalvariance2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Shieh's $d_s$, when variances are unequal across groups and sample sizes are equal, as a function of $\\sigma_1$, for a constant $SD$-ratio.",echo=FALSE}
par(mfrow=c(1,2))
plot(SD1,DF,xlab=expression(sigma[1]))
#SDratio[DF==max(DF)]
plot(SD1,round(coeffmult,6),ylab=expression(gamma),xlab=expression(sigma[1]))
#SDratio[coeffmult==min(coeffmult)]
```

Note: for a constant $SD$-ratio, $\sigma_1$ and $\sigma_2$ do not matter (see Figure \ref{fig:biasshiehhetbalvariance2}). 

### When variances are unequal across populations, with unequal sample sizes

The **bias** of Shieh's $d^*_s$ is a function of the population effect size ($\delta_{Shieh}$), the sample sizes ($n_1$ and $n_2$), and the interaction between the sample sizes ratio and the $SD$-ratio $\left(\frac{n_1}{n_2}\times\frac{\sigma_1}{\sigma_2} \right)$:

  + The larger the population effect size, the more Shieh's $d_s$ will overestimate $\delta_{Shieh}$;  

```{r biasshiehhetunbalNsize,include=FALSE}
coeffmult <- NULL
Nsize <- NULL
DF <- NULL

for (i in 2:200){
  
  n1 <- i
  n2 <- 12*i
  N <- n1+n2
  sd1 <- 1.8
  sd2 <- 1
  
  df <- (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF <- c(DF,df)
  Nsize <- c(Nsize,N)
  coeffmult <- c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}

```

```{r biasshiehhetunbalNsize2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Shieh's $d_s$, when variances and sample sizes are unequal across groups, as a function of the total sample size ($N$).",echo=FALSE}
par(mfrow=c(2,2))
plot(Nsize,DF,xlab="N")
plot(Nsize,coeffmult,ylab=expression(gamma),xlab="N")
```

  + For a constant sample sizes ratio, the larger the sample sizes, the lower the bias (See Figure \ref{fig:biasshiehhetunbalNsize2});

```{r biasshiehhetuneqNsize,include=FALSE}
Nsize2=NULL
coeffmult2=NULL
DF2=NULL

sd1=1.8
sd2=1

for (i in 4:200){
  n1=i
  n2=13
  N = n1+n2
  Nsize2=c(Nsize2,N)
  df = (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF2 = c(DF2,df)
  coeffmult2 = c(coeffmult2,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}

Nsize4=NULL
coeffmult4=NULL
DF4=NULL

sd1=1
sd2=1.8

for (i in 4:200){
  n1=i
  n2=13
  N = n1+n2
  Nsize4=c(Nsize4,N)
  df = (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF4 = c(DF4,df)
  coeffmult4 = c(coeffmult4,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}

Nsize=NULL
coeffmult=NULL
DF=NULL

sd1=1.8
sd2=1

for (i in 4:200){
  n1=13
  n2=i
  N = n1+n2
  Nsize=c(Nsize,N)
  df = (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF = c(DF,df)
  coeffmult = c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}


Nsize3=NULL
coeffmult3=NULL
DF3=NULL

sd1=1
sd2=1.8

for (i in 4:200){
  n1=13
  n2=i
  N = n1+n2
  Nsize3=c(Nsize3,N)
  df = (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF3 = c(DF3,df)
  coeffmult3 = c(coeffmult3,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}
```

```{r biasshiehhetuneqNsize2,fig.cap="Degrees of freedom (DF), when computing the bias of Shieh's $d_s$, when variances and sample sizes are unequal across groups, as a function of the total sample size, when adding subjects only in one group (either in the first group; see top plots; or in the second group; see bottom plots), and $\\sigma_1 > \\sigma_2$ (left plots) or $\\sigma_1 < \\sigma_2$ (right plots).",echo=FALSE}
par(mfrow=c(2,2))
plot(Nsize2,DF2,ylab="DF",xlab="N")
plot(Nsize4,DF4,ylab="DF",xlab="N")
plot(Nsize,DF,xlab="N")
plot(Nsize3,DF3,ylab="DF",xlab="N")
```

Note: When variances were equal across populations, adding subjects only in the first group had the same impact on degrees of freedom (and therefore on bias) than adding subjects only in the second group (see Figure \ref{fig:biasshiehhomuneqNsize2}). When variances are unequal across groups, this is not true anymore (see Figure \ref{fig:biasshiehhetuneqNsize2}).

```{r biasshiehhetunbalNratio,include=FALSE}
coeffmult2 <- NULL
nratio2 <- NULL
DF2 <- NULL
DF_NUM2 <- NULL
DF_DENOM2 <- NULL

for (i in 10:190){
  
  N <- 200
  n1 <- i
  n2 <- N-n1
  sd2 <- 1
  sd1 <- sqrt(50-sd2^2) # so sqrt((sd1^2+sd2^2)/2)=5
  SDratio2=sd1/sd2
  nratio2 <- c(nratio2,n1/n2)
  df_num <- (sd1^2/n1+sd2^2/n2)^2
  df_denom <- ((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  df <- (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF2 <- c(DF2,df)
  DF_NUM2 <- c(DF_NUM2, df_num)
  DF_DENOM2 <- c(DF_DENOM2, df_denom)

  coeffmult2 = c(coeffmult2,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}

coeffmult3 <- NULL
nratio3 <- NULL
DF3 <- NULL
DF_NUM3 <- NULL
DF_DENOM3 <- NULL

for (i in 10:190){
  
  N <- 200
  n1 <- i
  n2 <- N-n1
  sd2 <- 2
  sd1 <- sqrt(50-sd2^2) # so sqrt((sd1^2+sd2^2)/2)=5
  SDratio3=sd1/sd2
  nratio3 <- c(nratio3,n1/n2)
  df_num <- (sd1^2/n1+sd2^2/n2)^2
  df_denom <- ((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  df <- (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF_NUM3 <- c(DF_NUM3, df_num)
  DF_DENOM3 <- c(DF_DENOM3, df_denom)
  DF3 <- c(DF3,df)
  coeffmult3 = c(coeffmult3,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}

coeffmult4 <- NULL
nratio4 <- NULL
DF4 <- NULL
DF_NUM4 <- NULL
DF_DENOM4 <- NULL

for (i in 10:190){
  
  N <- 200
  n1 <- i
  n2 <- N-n1
  sd2 <- 4
  sd1 <- sqrt(50-sd2^2) # so sqrt((sd1^2+sd2^2)/2)=5
  SDratio4=sd1/sd2
  nratio4 <- c(nratio4,n1/n2)
  df_num <- (sd1^2/n1+sd2^2/n2)^2
  df_denom <- ((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  df <- (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF_NUM4 <- c(DF_NUM4, df_num)
  DF_DENOM4 <- c(DF_DENOM4, df_denom)
  df <- ((n1-1)*(n2-1)*(sd1^2+sd2^2)^2)/((n2-1)*sd1^4+(n1-1)*sd2^4)
  DF4 <- c(DF4,df)
  coeffmult4 = c(coeffmult4,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))
}
```

```{r biasshiehhetunbaldfandvar,fig.cap="Degrees of freedom ($DF$) and $\\gamma$, when computing the bias of Shieh's $d_s$, when variances and sample sizes are unequal across groups, as a function of the logarithm of the sample sizes ratio ($log \\left( \\frac{n_1}{n_2} \\right)$), when $SD$-ratio equals 1.46 (first row), 3.39 (second row) or 7 (third row).",echo=FALSE}
par(mfrow=c(3,2))

plot(log(nratio4),DF4,ylab="DF",xlab=expression(paste("log(",n[1],"/",n[2],")")))
plot(log(nratio4),coeffmult4,ylab=expression(gamma))

plot(log(nratio3),DF3,ylab="DF",xlab=expression(paste("log(",n[1],"/",n[2],")")))
abline(v=0)
plot(log(nratio3),coeffmult3,ylab=expression(gamma))
abline(v=0)

plot(log(nratio2),DF2,ylab="DF",xlab=expression(paste("log(",n[1],"/",n[2],")")))
abline(v=0)
plot(log(nratio2),coeffmult2,ylab=expression(gamma))
abline(v=0)

```

  + The smallest bias always occurs when there is a positive pairing between variances and sample sizes. Moreover, the further the $SD$-ratio is from 1, the further from 1 will also be the sample sizes ratio associated with the smallest bias (See Figure \ref{fig:biasshiehhetunbaldfandvar});

```{r biasshiehhetunbalvariance,include=FALSE}
coeffmult <- NULL
SD <- NULL
DF <- NULL

for (i in 2:200){
  
  n1 <- 60
  n2 <- 12
  N <- n1+n2
  sd1 <- i
  sd2 <- 8*i
  
  df <- (sd1^2/n1+sd2^2/n2)^2/((sd1^2/n1)^2/(n1-1)+(sd2^2/n2)^2/(n2-1))
  DF <- c(DF,df)
  SD <- c(SD,sqrt((sd1^2+sd2^2)/2))
  coeffmult <- c(coeffmult,sqrt(df/2)*gamma((df-1)/2)/gamma(df/2))   
}

```

```{r biasshiehhetunbalvariance2,fig.cap="Degrees of freedom (DF) and $\\gamma$, when computing the bias of Shieh's $d_s$, when variances and sample sizes are unequal across groups, as a function of $\\sigma_1$ and $\\sigma_2$, for a constant $SD$-ratio.",echo=FALSE}
par(mfrow=c(2,2))
plot(SD,DF,xlab=expression(sigma))
abline(v=0)
plot(SD,coeffmult,ylab=expression(gamma),xlab=expression(sigma))
abline(v=0)
```
  
Moreover, for a constant $SD$-ratio, the variances do not matter (See Figure \ref{fig:biasshiehhetunbalvariance2}).

## In summary

The **bias** of Cohen's $d_s$ is a function of the population effect size $\delta_{Cohen}$ and the total sample size ($N$): 

  + When $\delta_{Cohen}$ is null, the bias is null. In all other configurations, the larger $\delta_{Cohen}$, the more Cohen's $d_s$ will overestimate $\delta_{Cohen}$;  
  + The bias decreases when the total sample size increases (it does not matter whether one adds subjects in only one group or in both).  

The **bias** of Glass's $d_s$ is a function of the population effect size ($\delta_{Glass}$) and the size of the control group ($n_e$):  

  + When $\delta_{Glass}$ is null, the bias is null. In all other configurations, the larger $\delta_{Glass}$, the more Glass's $d_s$ will overestimate $\delta_{Glass}$;  
  + The bias decreases when the size of the control group increases. On the other hand, increasing the size of the experimental group does not impact the bias.  

The **bias** of Cohen's $d^*_s$ is a function of the population effect size ($\delta^*_{Cohen}$), the total sample size, and the interaction between the sample sizes ratio and the $SD$-ratio $\left(\frac{n_1}{n_2}\times\frac{\sigma_1}{\sigma_2} \right)$:       

  + When $\delta^*_{Cohen}$ is null, the bias is null. In all other configurations, the larger $\delta^*_{Cohen}$, the more Cohen's $d^*_s$ will overestimate $\delta^*_{Cohen}$;  
  + The bias decreases when the total sample size increases (it does not matter whether one adds subjects in only one group or in both);
  + The smallest bias always occurs when there is a positive pairing between $\frac{\sigma_1}{\sigma_1}$ and $\frac{n_1}{n_1}$. Moreover, the larger the $SD$-ratio, the further from 1 is the sample sizes ratio associated with the smallest bias.  

The **bias** of Shieh's $d_s$ is a function of the population effect size ($\delta_{Shieh}$), the total sample size, and the interaction between the sample sizes ratio and the $SD$-ratio $\left(\frac{n_1}{n_2}\times\frac{\sigma_1}{\sigma_2} \right)$:   

  + When $\delta_{Shieh}$ is null, the bias is null. In all other configurations, the larger $\delta_{Shieh}$, the more Shieh's $d_s$ will overestimate $\delta_{Shieh}$;    
  + For a constant sample sizes ratio, the bias decreases when the total sample size increases;  
  + The smallest bias always occurs when there is a positive pairing between $\frac{\sigma_1}{\sigma_1}$ and $\frac{n_1}{n_1}$. Moreover, the larger the $SD$-ratio, the further from 1 is the sample sizes ratio associated with the smallest bias (for more details, see "Theoretical Bias, as a function of population parameters").    
