---
title             : "Skewness and kurtosis: relation between Cain et al. (2017) and the package 'PearsonDS'"
shorttitle        : "G1 and G2"

author:
  - name          : "Marie Delacre"
    affiliation   : "1"
    corresponding : yes
    address       : "CP191, avenue F.D. Roosevelt 50, 1050 Bruxelles"
    email         : "marie.delacre@ulb.ac.be"

affiliation:
  - id            : "1"
    institution   : "Service of Analysis of the Data, Université Libre de Bruxelles, Belgium"

authornote: |

abstract: |

keywords          : "keywords"
wordcount         : "X"

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

In 2017, Cain et al. have conducted a review assessing the skewness and kurtosis of articles in recent psychology and education publications. They used the following formulas of Fisher's skewness ($G_{1}$) and kurtosis ($G_{2}$):

\begin{equation} 
G_{1}=\frac{\sqrt{n(n-1)}}{n-2} \frac{m_{3}}{\sqrt{(m_{2})^3}}
(\#eq:skew)
\end{equation} 

With n = sample size,  $m_{2}$ = second centered moment and $m_{3}$ = third centered moment.

\begin{equation} 
G_{2}=\frac{n-1}{(n-2)(n-3)}\times [(n+1)(\frac{m_{4}}{(m_{2})^2}-3)+6]
(\#eq:kurt)
\end{equation} 

with n = sample size and $m_{2}$ and $m_{4}$ = the second and fourth centered moments. I chose to use this article in order to define which value of skewness and kurtosis I would simulate, in order to test the goodness of different measures of effect sizes under realistic population parameter values. In my simulations, I Chose the function "rPearson" from the package "PearsonDS", in which skewness and kurtosis are computed as following: 

\begin{equation} 
skewness=\frac{m_{3}}{\sqrt{(m_{2})^3}}
(\#eq:skewnessrPearson)
\end{equation} 

\begin{equation} 
kurtosis=\frac{m_{4}}{(m_{2})^2}
(\#eq:kurtosisrPearson)
\end{equation} 

In order to simulate a sample extracted from a population where $G_{1}= X$, using the "rPearson" function, I need to make the following transformation:

\begin{equation} 
\frac{\sqrt{n(n-1)}}{n-2} \frac{m_{3}}{\sqrt{(m_{2})^3}} = X
<==> \frac{m_{3}}{\sqrt{(m_{2})^3}} = \frac{X(n-2)}{\sqrt{n(n-1)}}
(\#eq:skewnesstransformation)
\end{equation} 

In order to simulate a sample extracted from a population where $G_{2}= X$, using the "rPearson" function, I need to make the following transformation:

\begin{equation} 
\frac{n-1}{(n-2)(n-3)} [(n+1)(\frac{m_{4}}{(m_{2})^2}-3)+6] = X  
<==> \frac{m_{4}}{(m_{2})^2}=\frac{X(n-2)(n-3)-6(n-1)}{n^2-1}+3
(\#eq:kurtosistransformation)
\end{equation} 
