---
title             : "Reminder about Confidence Intervals"
shorttitle        : "CI REMINDER"

author: 
  - name          : "Marie Delacre"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "marie.delacre@ulb.ac.be"

affiliation:
  - id            : "1"
    institution   : "ULB"

authornote: |

abstract: |

keywords          : "keywords"
wordcount         : "X"

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
library("moments")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

## Reference

Cumming, G., & Finch, S. (2001). A primer on the understanding, use, and calculation of confidence intervales that  are based on central and noncentral distributions. Educational and Psychological Measurement, 61(532). 

# Method based on the use of a pivotal quantity

When computing a (supposed normal) centered variable, divided by the standard error (i.e. an independant variable closely related with the $\chi^2$ distribution), then computed quantity will follow a central *t*-distribution. This quantity is called a pivotal quantity (PQ), i.e. a quantity that is very interesting because its sampling distribution is not a function of the parameter we want to estimate (Cox & Hinkley, 1974 cited by Cumming and Finch, 2001). We can therefore use it, in order to define confidence limits for any parameter.

The method consists in four steps:  
1) Compute a pivotal quantity (PQ) of the general form: (Estimator - parameter)/SE;  
2) Determining the distribution of PQ;  
3) Computing the confidence limits of PQ: determine a range of values, centered around 0, such as (1-alpha)% of the area under the distribution of PQ falls in this range;    
4) Pivote in order to obtain the confidence interval around the parameter of interest.    

As a first example, consider the case of 2 means difference, assuming normality and homoscedasticity. The pivotal quantity is defined as follows: 

\begin{equation} 
PQ= \frac{(\bar{X_1}-\bar{X_2})-(\mu_1-\mu_2)}{SE}
(\#eq:PQstudent)
\end{equation} 

With $SE = \sigma_{pooled} \times \sqrt{\frac{1}{n_1}+\frac{1}{n_2}}$ and $\sigma_{pooled} = \sqrt{\frac{(n_1-1)*S^2_1+(n_2-1)*S^2_2}{n_1+n_2-2}}$

This quantity follows a *t*- distribution with $n_1+n_2-2$ degrees of freedom (therefore, it depends only on $n_1$ and $n_2$, it does NOT depend on the parameter of interest, i.e. $\mu_1-\mu_2$).

```{r, SAMPLMEANDIFF1, fig.cap= "Sampling distribution of the pivotal quantity under the assumptions of normality and homoscedasticity"}
samplingtest <- function(n1,n2,nsim){
  
  for (i in seq_len(nsim)){
    A <- rnorm(n1,mean=4)
    B <- rnorm(n2,mean=2)
    if (i==1){
      av <- mean(A)-mean(B)  
      sd1 <- sd(A)
      sd2 <- sd(B)
      
    } else {
      av <- c(av,mean(A)-mean(B))
      sd1 <- c(sd1,sd(A))
      sd2 <- c(sd2,sd(B))
    }
    
  }
  
  pooled_sd <- sqrt(((n1-1)*sd1^2+(n2-1)*sd2^2)/(n1+n2-2))
  pq <- (av-(4-2))/(pooled_sd*sqrt(1/n1+1/n2))
  
  1-pt(.025,df=n1+n2-2)
  plot(density(pq),main="Sampling distribution of PQ",xlab="",cex.main=1.5,ylim=c(0,.5)) # plotting the sampling distribution of the pivotal quantity
  legend("top",col=c("black","red"),legend=c("empirical distribution (based on simulations)","t-distribution with n1+n2-2 degrees of freedom"),lty=c(2,2),bty="n")
  
  df <- n1+n2-2
  sq <- seq(-4,4,.01) 
  lines(sq,dt(sq,df=df),lty=2,pch=9,cex=.5,col="red")
  
}

samplingtest(n1=50,n2=60,nsim=100000)

```

Because the theoretical distribution of PQ is known, one can compute the confidence limits, for any confidence level:

\begin{equation} 
Pr[t_{n_1+n_2-2}(\frac{\alpha}{2}) < \frac{(\bar{X_1}-\bar{X_2})-(\mu_1-\mu_2)}{SE} < t_{n_1+n_2-2}(1-\frac{\alpha}{2})] = 1 - \alpha
(\#eq:conflev1)
\end{equation} 

Because the *t*-distribution is symmetrically centered around 0, one can deduce that $t_{n_1+n_2-2}(\frac{\alpha}{2})=-t_{n_1+n_2-2}(1-\frac{\alpha}{2})$, and therefore:

\begin{equation} 
Pr[-t_{n_1+n_2-2}(1-\frac{\alpha}{2}) < \frac{(\bar{X_1}-\bar{X_2})-(\mu_1-\mu_2)}{SE} < t_{n_1+n_2-2}(1-\frac{\alpha}{2})] = 1 - \alpha
(\#eq:conflev2)
\end{equation} 

In pivoting the inequation, one can deduce that:

\begin{equation} 
Pr[-t_{n_1+n_2-2}(1-\frac{\alpha}{2}) \times SE < (\bar{X_1}-\bar{X_2})-(\mu_1-\mu_2) < t_{n_1+n_2-2}(1-\frac{\alpha}{2}) \times SE] = 1-\alpha
(\#eq:conflev3)
\end{equation} 

\begin{equation} 
\leftrightarrow Pr[-(\bar{X_1}-\bar{X_2}) -t_{n_1+n_2-2}(1-\frac{\alpha}{2}) \times SE <
-(\mu_1-\mu_2) 
< -(\bar{X_1}-\bar{X_2}) +t_{n_1+n_2-2}(1-\frac{\alpha}{2}) \times SE]= 1- \alpha
(\#eq:conflev4)
\end{equation} 

\begin{equation} 
\leftrightarrow Pr[(\bar{X_1}-\bar{X_2}) +t_{n_1+n_2-2}(1-\frac{\alpha}{2}) \times SE > \mu_1-\mu_2 > (\bar{X_1}-\bar{X_2}) - t_{n_1+n_2-2}(1-\frac{\alpha}{2}) \times SE]= 1- \alpha
(\#eq:conflev5)
\end{equation} 

\begin{equation} 
\leftrightarrow Pr[(\bar{X_1}-\bar{X_2}) - t_{n_1+n_2-2}(1-\frac{\alpha}{2}) \times SE < \mu_1-\mu_2 <(\bar{X_1}-\bar{X_2}) +t_{n_1+n_2-2}(1-\frac{\alpha}{2}) \times SE]= 1- \alpha
(\#eq:conflev6)
\end{equation} 

As a second example, consider the case of 2 means difference, assuming normality and heteroscedasticity. The pivotal quantity is defined as follows:

\begin{equation} 
PQ= \frac{(\bar{X_1}-\bar{X_2})-(\mu_1-\mu_2)}{SE}
(\#eq:PQwelch)
\end{equation} 

With $SE = \sqrt{\frac{S^2_1}{n1}+\frac{S^2_2}{n2}}$

This quantity follows a *t*- distribution with $\frac{(\frac{S^2_1}{n_1}+\frac{S^2_2}{n_2})^2}{\frac{(\frac{S^2_1}{n_1})^2}{n_1-1}+\frac{(\frac{S^2_2}{n_2})^2}{n_2-1}}$ degrees of freedom (therefore, it depends on $n_1$ and $n_2$, $S_1$ and $S_2$, and does NOT depend on the parameter of interest, i.e. $\mu_1-\mu_2$).

```{r, SAMPLMEANDIFF2, fig.cap= "Sampling distribution of the pivotal quantity under the assumptions of normality and heteroscedasticity"}
samplingtest2 <- function(n1,n2,sigma1,sigma2,nsim){
  
  for (i in seq_len(nsim)){
    A <- rnorm(n1,mean=4,sd=sigma1)
    B <- rnorm(n2,mean=2,sd=sigma2) 
    if (i==1){
      av <- mean(A)-mean(B)  
      sd1 <- sd(A)
      sd2 <- sd(B)
      
    } else { 
      av <- c(av,mean(A)-mean(B))
      sd1 <- c(sd1,sd(A))
      sd2 <- c(sd2,sd(B))
    }
    
  }
  
  pq <- (av-(4-2))/sqrt(sd1^2/n1+sd2^2/n2) 
  plot(density(pq),main="Sampling distribution of PQ",xlab="",cex.main=1.5,ylim=c(0,.5)) # plotting the sampling distribution of the pivotal quantity
  legend("top",col=c("black","red"),legend=c("empirical distribution (based on simulations)","t-distribution with df degrees of freedom"),lty=c(2,2),bty="n")

  df <- (sigma1^2/n1+sigma2^2/n2)^2/(((sigma1^2/n1)^2)/(n1-1)+((sigma2^2/n2)^2)/(n2-1)) 
  sq <- seq(-4,4,.01) 
  lines(sq,dt(sq,df=df),lty=2,pch=9,cex=.5,col="red")
  
}

samplingtest2(n1=50,n2=60, sigma1=2,sigma2=5,nsim=100000)
```

Because the theoretical distribution of PQ is known, one can compute the confidence limits, for any confidence level (see the first example for more details):

\begin{equation} 
Pr[(\bar{X_1}-\bar{X_2}) - t_{DF}(1-\frac{\alpha}{2}) \times SE < \mu_1-\mu_2 <(\bar{X_1}-\bar{X_2}) +t_{DF}(1-\frac{\alpha}{2}) \times SE]= 1- \alpha
(\#eq:conflev6)
\end{equation} 

With SE = $\sqrt{\frac{S_1^2}{n_1}+\frac{S_2^2}{n_2}}$



